---
title: "2. Visualizations using caret"
author: "Russ Conte"
date: "4/29/2022"
output: html_document
---

The `featurePlot` function is a wrappter for different `lattice` plots to visualize the data. For example, the following figures show the default plot for continuous outcomes generated using the `featurePlot` function.

For classification data sets, the <span style="color:blue">`iris`</span> data are used for illustration.

```{r information about the `iris` data set}
str(iris)
```

### Scatterplot Matrix

This first block of code loads all the required packages in two lines - nice and clean!

```{r load all packages from two lines of code}
Packages = c("AppliedPredictiveModeling", "caret", "tidyverse", "mlbench")
lapply(Packages, library, character.only = TRUE)
```

Let's create the scatterplot matrix for the iris data set!

```{r create a scatterplot matrix of the data in the iris data set}
transparentTheme(trans = 0.4)
featurePlot(x = iris[,1:4],
            y = iris$Species,
            plot = "pairs",
            ## Add a key at the top to list the names of the irises
            auto.key = list(columns = 3))

```

### Scatterplot Matrix with Ellipses

```{r Scatterplot Matrix withEllipses}

featurePlot(x = iris[,1:4],
             y = iris$Species,
            plot = "ellipse",
            ## Add a key at the top of the data visualization
            auto.key = list(columns = 3))

# A working example of the same solution using the mpg data set from tidyverse:

mpg1 <- mpg %>% 
  mutate(class = as.factor(class)) # convert y-hat to a factor

featurePlot(x = mpg1[,c(3,8:9)],
            y = mpg1$class,
            plot = "ellipse",
            auto.key = list(7))

```

### Overlayed Density Plots (I like these a lot!)

```{r Overlayed Density Plots}

transparentTheme(trans = 0.9)
featurePlot(x = iris[1:4],
            y = iris$Species,
            plot = "density")

```

```{Repeat density plots, but pass in options to xyplot() to make it prettier}

featurePlot(x = iris[,1:4],
            y = iris$Species,
            plot = "density",
            scales = list(x = list(relation = "free"),
                          y = list(relation = "free")),
            adjust = 1.5,
            pch = "|",
            layout = c(4,1),
            auto.key = list(columns = 3))


```

Repeat similar exercise for density plots using mpg data set from tidyverse:

```{r Repeat density plot exercise with mpg data from tidyverse}

featurePlot(x = mpg1[,c(3,8:9)],
            y = mpg1$class,
            plot = "density",
            scales = list(x = list(relation = "free"),
                          y = list(relation = "free"),
            adjust = 1.5),
            pch = "|",
            auto.key = list(columns = 3)) # creates three columns for the labels - nice!

```

### Box plots

```{r Box plots using caret and the iris data set}

featurePlot(x = iris[,1:4],
            y = iris$Species,
            plot = "box",
# pass in options to bwplot()
            scales = list(y = list(relation = "free"),
                          x = list(rot = 90)),
            layout = c(4,1),
            auto.key = list(columns = 2))

```

Repeat similar exercise for box plots using mpg data from tidyverse:

```{r Box plots using caret and the mpg data set from tidyverse}

featurePlot(x = mpg1[,c(3, 8:9)],
            y = mpg1$class,
            plot = "box",
            scales = list(y = list(relation = "free"),
                          x = list(rot = 90)),
            layout = c(4,1),
            auto.key = list(columns = 2))

```


### Scatter Plots

For regression, we'll start with the Boston housing data set:

```{r First regression exammple}
data("BostonHousing")
regVar <- c("age", "lstat", "tax")
str(BostonHousing[, regVar])

```

When the predictors are continuous, `featurePlot` can be used to create scatter plots of each of the predictors with the outcome. For example:

```{r Scatterplots with continuous predictors}

theme1 <- trellis.par.get()
theme1$plot.symbol$col = rgb(.2, .2, .2, .4)
theme1$plot.symbol.pch = 16
theme1$plot.line$col = rgb(1,0,0,.7)
theme1$plot.line$lwd <- 2
trellis.par.set(theme1)
featurePlot(x = BostonHousing[,regVar],
            y = BostonHousing$medv,
            plot = "scatter",
            layout = c(3,1))

```

Note that the x-axis scales are different from one another. The function automaticallyuses `scales = list(y = list(relation = "free`)) so you don't have to add it. We can also pass in options to the `lattice` function `xyplot`. For example, we can add a scatter plot smoother by passing in new options:

```{r}
featurePlot(x = BostonHousing[, regVar],
            y = BostonHousing$medv,
            plot = "scatter",
            type = c("p","smooth"),
            span = 0.5,
            layout = c(3,1))
```

Note that the options `degree` and `span` control the smoothness of the smoother.

Let's create a feature plot with the continuous variables in the mpg data set. Let's predict and plot highway mileage given the four other continuous values in the data set:

```{r Create scatter plots in the mpg data set from tidyverse}

featurePlot(x = mpg1[,c(3:5,8)],
            y = mpg1$hwy,
            plot = "scatter",
            type = c("p", "smooth"),
            span = 0.5,
            layout = c(4,1))


```

Summary of the graph: Clearly the city mileage is the strongest predictor of the four for highway mileage.

This completes chapter 2, Visualizations, it was a lot of fun, and I learned a lot (as always!)