---
title: "3. Pre-Processing using caret"
author: "Russ Conte"
date: "4/30/2022"
output: html_document
---

`caret` includes several functions to pre-process the predictor data. It assumes that all of the data are numeric (i.e. factors have been converted to dummy variables via `model.matrix`, `dummyVars`, or other means.)

### 3.1 Creating Dummy Variables

The function `dummuVars` can be used to generate a complete (less than full rank parameterized) set of dummy variables from one or more factors. The function takes a formula and a data set and outputs an object that can be used to create the dummy variables using the predict method.

For example, the `etitanic` data set in the <span style="color:blue">`earth`</span> package includes two factors: `pclass` (passenger class, with levels 1st, 2nd, and 3rd) and `sex` (with levels female, male).

Let's load all the packages in two lines of code:

```{r}
Packages = c("AppliedPredictiveModeling", "caret", "tidyverse", "e1071", "earth")
lapply(Packages, library, character.only = TRUE)
```

Using `dummyVars`:

```{r Using dummyVars to create dummy variables in the titanic data set}

dummies <- dummyVars(survived~., data = etitanic)
head(predict(dummies, newdata = etitanic))

```

The authors point out there is no intercept, and each factor has a dummy variable for each level. This may not be useful for some modeling situations, such as `lm`.

