---
title: "5. Model Training and Tuning"
author: "Russ Conte"
date: '2022-05-22'
output: html_document
---

# Model Training and Tuning

Contents

[5.1 Model Training and Parameter Tuning](#Model Training and Parameter Tuning)<br>
[5.2 An Example](#5.2 An Example)<br>
[5.3 Basic Parameter Tuning](#5.3 Basic Parameter Tuning)<br>
[5.4 Notes on Reproducibility](#5.4 Notes on Reproducibility)<br>
[5.5 Customizing the Tuning Process](#5.5 Customizing the Tuning Process)<br>
[5.5.1 Pre-Processing Options](#5.5.1 Pre-Processing Option)<br>
[5.5.2 Alternate Tuning Grids](#Alternate Tuning Grids)<br>
[5.5.3 Plotting the Resampling Profile](#Plotting the Resampling Profile)<br>
[5.5.4 The `trainControl` Function](#5.5.4 The `trainControl` Function)<br>
[5.5.5 Alternate Performance Metrics](#Alkternate Performance Metrics)<br>
[5.6 Choosing the Final Model](#Choosing the Final Model)<br>
[5.7 Extracting Predictions and Class Probabilities](#Extracting Predictions and Class Probabilities)<br>
[5.8 Exploring and Comparing Resampling Distrubutions](#Exploring and Comparing Resampling Distributions)<br>
[5.8.1 Within-Model](#5.8.1 Within-Model)<br>
[5.8.2 Between-Models](#Between-Models)<br>
[5.9 Fitting Models Without Parameter Tuning](#5.9 Fitting Models Without Parameter Tuning)<br>


### <a id="5.1 Model Training and Parameter Tuning"></a>5.1 Model Training and Paramter Tuning###

The `caret` package has several functions that attempt to streamline the model building and evaluaton process.

The `train` function can be used to:
* Evaluate, using resampling, the effect of model tuning parameters on model performance
* Choose the "optimal" model across these parameters
* Estimate model performance from a training set

First, a specific model must be chosen. Current, 238 are available using `caret`.

The first step in tuning the model (line 1 in the algorithm below) is to choose a set of parameters to evaluate. For example, if fitting a Partial Least Squares (PLS) model, the number of PLS components to evaluate must be specified:

1. Define sets of model parameter values to evaluate
2. **for** *each parameter set* **do**
3.    **for** *each resampling iteration* **do**<br>
    4. Hold-out Specific Items<br>
    5. [Optional] Pre-Process the data<br>
    6. Fit the model on the remainder<br>
    7. Predict the Hold-out samples<br>
8.    **end**
9.    Calculate the average performanc across hold-out predictions
10. **end**
11. Determine the optimal parameter set
12. Fit the final model to all the training data using the optimal parameter set.


Once the model and tuning parameter values have been defined, the type of resampling should also be specified. Currently, *k*-fold cross-validation(once or repeated), leave-one-out cross-validation and bootstrap (simple estimation or the 632 rule) resampling methods can be used by `train`. After resampling, the process produces a profile of performance measure available to guide the user as to which tuning parameter values should be chosen. By default, the function augomatically chooses the tuning parameters associationed with the best values, although different algorithms can be used (see details below)


### <a id="5.2. An Example"></a>5.2. An Example###

The Sonar data are available in the `mlbench` package. Here, we load the data:

```{r An example, using the Sonar data}

library(mlbench)
data(Sonar)
str(Sonar[,1:10])

```

The function `createDataPartition` can be used to create a stratified random sample of data into training and test sets:

```{r Create stratified random samples of the data into training and test sets}

library(caret)
set.seed(998)
inTraining <- createDataPartition(Sonar$Class, p = 0.75, list = FALSE)
training <- Sonar[inTraining,]
testing <- Sonar[-inTraining,]

training[1:10, 1:10] # 157 observations of 61 variables
testing[1:10, 1:10] # 51 observations of 61 variables

```


### <a id="5.3 Basic Parameter Tuning"></a>5.3 Basic Parameter Tuning###

By default, simple bootstrap resampling is used for line 3 in the algorithm above. Other methods are available, such as *K*-fold cross-validation, leave-one-out, etc. The function, `trainControl` can be used to specify the type of resampling:

```{r Using trainControl to set up resampling}

fitControl <- trainControl(## 10-fold cross-validation,
                            method = "repeatedcv",
                            number = 10,
                            ## repeated ten times,
                            repeats = 10)

```


More information about `trainControl` is given in a section below.

The first two arguments to `train` are the predictor and outcome data objects, respectively. The third argument, `method`, specifies the type of model. To illustrate, we will fit a boosted tree model via the `gbm` package. The basic syntax for fitting this model using repeated cross-validation is shown below:

```{r Fitting a boosted tree using gbm}

set.seed(825)
gbmfit1 <- train(Class~., data = training,
                 method = "gbm",
                 trControl = fitControl,
                 ## This last option is actually one for gbm() that passes through
                 verbose = FALSE)
gbmfit1

```

For a gradient boosting machine (GBM) model, there are three (?) main tuning parameters:

* number of iterations, i.e. trees, (called `n.trees` in the `gbm` function)
* complexity of the tree, called `interaction.depth`
* learning rate: how quickly the algorithm adapts, called `shrinkage`
* the minimum number of training set samples in a note to commence splitting(`n.minobsinnode`)

The default values tested for this model are shown in the first two columns (`shrinkage` and `n.minobsinnode` are not shown because the grid set of candidate models all use a single value for these tuning parameters). The column labeled `accuracy` is the overall agreement rate averaged over cross-validation iterations. The agreement standard deviation is also calculated from the cross-validation results. The column, `Kappa` is Cohen's (unweighted) Kappa statistic averaged across the resampling results. `train` works with specific models. For these models, `train` can automatically create a grid of tuning parameters. By default, if *p* is the number of tuning parameters, the grid size is 3^p. As another example, regularized discriminant analysis (RDA) models have two parameters (`gamma` and `lambda`), both of which lie between zero and one. The default trainng grid would produce nine combinations in this two-dimensional space.

### <a id="5.4 Notes on Reproducibility"></a>5.4 Notes on Reproducibility###

Many models utilize random numbers during the phase where parameters are estimated. Also, the resampling indices are chosen using random numbers. There are two main ways to control the randomness in order to assure reproducible results:

* There are two approaches to ensuring that the same *resamples* are used between calls to `train`. The first is to use `set.seed` just prior to calling `train`. The first use of random numbers is to create the resampling information. Alternatively, if you would like to use specific splits of the data, the `index` argument of the `trainControl` function can be used. This is briefly discussed below.
* When the models are created *inside of resampling*, the seeds can also be set. While setting the seed prior to calling `train` may guarantee that the same rendom numbers are used, this is unlikely to be the case when parallel processing is used (depending on which technology is utilized). To set the model fitting seeds, `trainControl` has an additional argument called `seeds` that can be used. The value for this argument is a list of integer vectors that are used as seeds. The help oage for `trainControl` describes the appropriate format for this option.



### <a id="5.5 Customizing the Tuning Process"></a>5.5 Customizing the Tuning Process###


#### <a id="5.5.1 Pre-Processing Options"></a>5.5.1 Pre-Processing Options###



#### <a id="5.5.2 Alternate Tuning Grids"></a>5.5.2 Alternate Tuning Grids###


#### <a id="5.5.3 Plotting the Resampling Profile"></a>5.5.3 Plotting the Resampling Profile###


#### <a id="5.5.4 The `trainControl` Function"></a>5.5.4 The `trainControl` Function###


#### <a id="5.5.5 Alternate Performance Metrics"></a>5.5.5 Alternate Performance Metrics###


### <a id="5.6 Choosing the Final Model"></a>5.6 Choosing the Final Model###


### <a id="5.7 Extracting Predictions and Class Probabilities"></a>5.7 Extracting Predictions and Class Probabilities###



### <a id="5.8.2 Between-Models"></a>5.8.2 Between-Models###



### <a id="5.9 Fitting Models Without Parameter Tuning"></a>Fitting Models Without Parameter Tuning###


